The Talking Heads Experiment

### 2015-04-16 chap6/chap6.tex
hyphenation, orphans, widows and some other layout updated

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape obj} & HPOS & VPOS & HEIGHT & WIDTH & GRAY & AREA \\ \midrule
0 & 0.66 & 0.95 & 0.01 & 0.71 & 0.19 & 0.27\\ 
1 & 0.69 & 0.83 & 0.07 & 0.33 & 0.97 & 0.21\\ 
2 & 0.99 & 0.87 & 0.54 & 0.72 & 0.22 & 0.57\\ 
saliency & 0.03 & 0.05 & 0.07 & 0.39 & 0.75 & 0.06 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-rect1}Sensory data for the scene shown in \figref{rect1}.}
\end{center}
\end{table}
All rectangles are relatively close to each other and have more or less the 
same height and width. But the grayscale is clearly the more salient because rectangle-1 is 
much darker than the others. I assume that there are only two agents in the population and that they always use only 
the most salient to conceptualise the scene. 
The speaker and hearer have to traverse only 
two sides of the semiotic square (\figref{square6})
because we assume that perceived image and object being
referred to are identical for both agents.

@@

The speaker first plays a Discrimination Game traversing the 
semantic side of the square going from the referent 
{\itshape rectangle-1} to a possible meaning [GRAY 0.5–1.0]. 
He then plays a Naming Game traversing the lexical side of 
of the square to find the word `pokuneso' for this chosen meaning. 
The hearer traverses the lexical side of the triangle in 
the other direction to interpret the word `pokuneso' as
{}[GRAY 0.5–1.0], and then identifies the referent by 
filtering the objects in the context with this meaning. 
Only rectangle-1 remains, so the game succeeds. 
The whole game is reported by the commentator as follows: 
\begin{verbatim}
Game 500
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a2 chooses rectangle-1 as the topic 
  a2 categorises the topic as [GRAY 0.5–1.0]
  a2 says: pokuneso
  a1 interprets pokuneso as [GRAY 0.5–1.0]
  a1 points to rectangle-1
  a2 signals OK 
\end{verbatim}
The game is perfectly successful because both 
agents associate the word `okuneso' with 
{}[GRAY 0.5–1.0] (dark) and they perceive the scene 
in the same way.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\itshape Translation} & {\bfshape  a1}&{\bfshape  a2} \\ \midrule
{}[HPOS 0.0–0.5] & vapola&left&-&0.1\\ 
{}[HPOS 0.5–1.0]& gonapa&right &0.1&-\\ 
{}[HEIGHT 0.0–0.5]&suwaxugo&short &0.6&0.8\\ 
{}[HEIGHT 0.5–1.0]& kusone&tall &0.4&0.5\\ 
{}[WIDTH 0.0–0.5]&bepupepa&narrow &0.1&0.1\\ 
{}[WIDTH 0.0–0.25]&kutaki&very narrow &-&0.1\\ 
{}[WIDTH 0.5–1.0]& zikorika&wide &0.0&0.3\\ 
{}[GRAY 0.0–0.5]& fesasado&light &0.5&0.7\\ 
{}[GRAY 0.5–1.0]& pokuneso&dark &0.8&0.9\\ 
{}[AREA 0.5–1.0]& mafanoda&large &0.1&0.1\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex100}. Complete lexicon of {\bfshape  a1} and {\bfshape  a2} after 100 games.}
\end{center}
\end{table}
We see that, at this point, the agents have lexicalised 
only the most general distinctions, such as dark (`pokuneso') 
versus light (`fesasado') or short (`suwaxugo') versus tall 
(`kusone'). Words for the grayscale and height dimensions
have the strongest scores, although this is purely accidental. 
When we would start another simulation from scratch,
we would end up with different words and perhaps 
other distinctions would be more successful.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\itshape Translation} & {\bfshape  a1}&{\bfshape  a2} \\ \midrule
{}[HPOS 0.0–0.5]&vapola&left &0.7&1.0\\ 
{}[HPOS 0.5–1.0]&gonapa&right &0.6&0.5\\ 
{}[VPOS 0.0–0.5]&rixuzime& up & 0.2&0.7\\ 
{}[VPOS 0.5–1.0]&gofugage& down &0.6&1.0\\ 
{}[HEIGHT 0.0–0.5]&suwaxugo&short & 1.0&1.0\\ 
{}[HEIGHT 0.0–0.25]&tawube&very short & 0.4&0.5\\ 
{}[HEIGHT 0.25–0.5]&narofi&medium short&0.1&0.4\\ 
{}[HEIGHT 0.5–1.0]&kusone&tall&1.0&1.0\\ 
{}[HEIGHT 0.5–0.75]&wuruzo&medium tall&0.3&0.6\\ 
{}[HEIGHT 0.75–1.0]&bowaluro&very tall&0.6&0.2\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex500a}Lexicon of {\bfshape  a1} and {\bfshape  a2} after 500 games.}
\end{center}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{}[WIDTH 0.0–0.5]&bepupepa&narrow & 1.0&1.0\\ 
{}[WIDTH 0.0–0.25]&kutaki&very narrow & 0.1&0.5\\ 
{}[WIDTH 0.25–0.5]&wukogo&medium narrow & 0.2&-\\ 
{}[WIDTH 0.5–1.0]&zikorika&wide & 1.0&1.0\\ 
{}[WIDTH 0.5–0.75]&mitula&medium wide &0.1&-\\ 
{}[WIDTH 0.75–1.0]&wupixo&very wide & -&0.2\\ 
{}[GRAY 0.0–0.5]&fesasado&light & 1.0&1.0\\ 
{}[GRAY 0.0–0.25]&sanize&very light & -&0.1\\ 
{}[GRAY 0.5–1.0]&pokuneso&dark &0.9&1.0\\ 
{}[GRAY 0.5–0.75]&wavosoru&medium dark & 0.2&0.5\\ 
{}[GRAY 0.75–1.0]&kuragoni&very dark &0.3&0.2\\ 
{}[AREA 0.0–0.5]&babifewa&small & -&0.1\\ 
{}[AREA 0.25–0.5]&togule&medium small & 0.1&0.1\\ 
{}[AREA 0.5–1.0]&mafanoda&large & 0.2&0.5\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex500b}Lexicon of {\bfshape  a1} and {\bfshape  a2} after 1000 games.}
\end{center}
\end{table}

We see that words for the basic distinctions have further established
themselves. Words for short (`suwaxugo') and tall (`kusone'), or light (`fesasado') and dark 
(`pokuneso'), now have scores of 1.0. Words for more refined categories, like very short (`tawube') or 
very narrow (`kutaki'), are beginning to establish themselves.

Two steps in the evolution of the discrimination trees
underlying this lexicon are shown in \figref{gdis1}.
There is a progressive refinement 
of all trees as time goes on, 
because all sensory channels have the same chance
of being most salient. But the trees are not 
the same for the two agents at every stage
of development because even though they prefer to expand
the salient channel, the agents have encountered 
different environmental situations in which different
channels were salient. For example, after 100 
games, {\bfshape  a1} has less refinements for the 
WIDTH channel than {\bfshape  a2}. 
After 500 games, all trees 
have at least one level of refinement. Not all 
categories have been lexicalised. For example, the 
WIDTH channel is three levels deep in both 
agents but no words exist yet for the deepest 
level.

@@

Also for the hearer, the re-entrant flow is important. 
A hearer can only know which meaning was intended for a particular
word after trying out the meaning on the scene. He
therefore uses the context to determine the meaning of 
the utterance. For example, a particular word may mean both [LARGE]
and [DARK], particularly during the phase of 
early language acquisition. However if only one of these categories
picks out a single referent, it is chosen as 
the meaning, and the hearer will act upon this choice by 
pointing to the object it singles out from the scene.

@@

The next thing I had to do is reconsider the score
updating mechanisms, even though they are basically 
the same as used earlier (\figref{incr-decr2}). 
For a given referent, there are multiple meanings possible
(in case more than one channel is considered to 
be sufficiently salient), and
for each meaning there are multiple words. The best 
one of this whole lot is chosen by the speaker and used
for the utterance transmitted to the hearer. We have 
seen that it is important for the hearer to use lateral
inhibition based on the outcome of a game. But although 
the speaker is considering {\itshape all} the possible 
conceptualisations, lateral
inhibition should only take place between the 
lexicalisations of the meaning that was finally chosen. 
So when the game was successful, 
the speaker increases the association that was used with $\delta$
and decreases all the other associations
{\itshape with the same meaning}. $delta$ is still set to a reasonable
low value, namely $delta = 0.1$. 
The hearer does the same as explained earlier in 
the Naming Game. The score of all
the alternative meanings for the word (or words) 
that are used by the speaker are decreased. 
When the game fails, both the speaker and the hearer 
decrease the association that they used with $\delta$. No
change takes place to the scores of any of the 
other associations.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap6/figs/incr-decr2.pdf}}
\caption{\label{incr-decr2}Score adjustments after a successful game. Used 
associations go up and competing associations go down.
The game producing these relationships is discussed
later as game 10008.}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj} & Hpos & Vpos & Height & Width & Gray & Area \\ \midrule
0 & 0.51 & 0.98 & 0.90 & 0.47 & 0.79 & 0.60\\ 
1 & 0.75 & 0.68 & 0.26 & 0.09 & 0.14 & 0.20\\ 
2 & 0.76 & 0.54 & 0.56 & 0.26 & 0.94 & 0.36\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game77}Sensory data for game 77 after sensor-scaling.}
\end{center}
\end{table}
The speaker is again {\bfshape  a2}. HEIGHT is the most salient channel. 
{\bfshape  a2} has a word for [HEIGHT 0.75–1.0] (very tall), namely
`bowaluro', and uses it. The hearer {\bfshape  a1} 
does not know the 
word, conceptualises the scene, and arrives at the 
same category, so the hearer stores the new word
with the same meaning as the speaker. 
\begin{verbatim}
Game 77
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a2 chooses rectangle-1 as the topic 
  a2 categorises the topic as [HEIGHT 0.75–1.0]
  a2 says: `bowaluro'
  a1 does not know `bowaluro'
  a1 says: `bowaluro?'
  a2 points to rectangle-1
  a1 categorises the topic as [HEIGHT 0.75–1.0]
  a1 stores `bowaluro' as [HEIGHT 0.75–1.0]
\end{verbatim}
The reason why {\bfshape  a1} has guessed the right 
meaning of `bowaluro' is because both agents use
only the most salient channel and they both share
the same perception of reality. We will soon see 
that if these constraints are not valid, agents 
are not always so lucky and hence multiple meanings
start to circulate for the same word.

@@

Part of the lexicons of five of the ten agents after 2000 games 
(only those words which have positive scores) are shown in the 
\tabref{tab:lex2000}. We see that synonymy does indeed occur. For
example, two words are in the 
running for [HPOS 0.5–1.0]: `rutaxese' and `xomupovi', and 
three for [VPOS 0.5–1.0]: `wavone', `zaxawe', and 
`dazofo'. Some words are already well established, for
example `numefuli' for [GRAY 0.5–1.0] (dark).

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l  }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\bfshape  a1}&{\bfshape  a2}&{\bfshape  a3}&{\bfshape  a4}&{\bfshape  a5} \\ \midrule
{}[HPOS 0.5–1.0]&rutaxese& &0.1& & &\\ 
 & xomupovi&0.1& &0.2& &0.1\\ 
{}[VPOS 0.5–1.0]&wavone& & &0.3& &\\ 
 & zaxawe& & & &0.1& \\ 
 & dazofo&0.1& & &0.2&\\ 
{}[WIDTH 0.0–0.5]&buxevo& & & & & \\ 
 & vubupo&1.0&0.6&1.0&0.1&1.0\\ 
{}[WIDTH 0.5–1.0]&pawixona& & &0.1& & \\ 
 & rikepule&1.0&0.8&1.0&1.0&1.0\\ 
{}[WIDTH 0.5–0.75]&gowinoge& & &0.1& &  \\ 
 & wesurodi& & & &0.2&\\ 
{}[WIDTH 0.75–1.0]&besabi& & &0.2& & \\ 
 & lituvi&0.1& & & & \\ 
{}[GRAY 0.5–0.75]&goxomixe& &0.1& & & \\ 
 & korufo& & & &0.2&\\ 
{}[GRAY 0.0–0.5]&numefuli&1.0&0.7&1.0&1.0&1.0\\ 
{}[GRAY 0.0–0.25]&rekemaxi&0.2& & & & \\ 
{}[GRAY 0.5–1.0]&faluleru&0.6&0.6&1.0&1.0&1.0\\ 
 & nupanu& & & & &0.2 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex2000}Lexicon of five agents after 2000 games.}
\end{center}
\end{table}

The positive feedback loop between use and success
has already dampened some synonyms. For some cases, like [GRAY 0.5–1.0] the competition has died out 
with one word `faluleru' being the winner. 
For others, like [HPOS 0.5–1.0], the competition 
is still going on, although we can guess
that `xomupovi' is probably going to be the winner.

It is instructive to follow the history of the words 
in use for a particular meaning. For 
example, let us look at the words for [GRAY 0.5–1.0] 
(dark) in the very early phases of lexicon development.
Three different words are quickly created, and 
`faluleru' is the first one that has some success. 
\begin{verbatim}
Game 76
 Speaker a4 creates `notabefe' for [GRAY 0.5–1.0]
 Hearer a5 adopts `notabefe' for [GRAY 0.5–1.0]
Game 79 
 Speaker a2 creates `vivevobo' for [GRAY 0.5–1.0]
 Hearer a7 does not adopt `vivebo' 
     (failed to discriminate)
Game 88 
 Speaker a7 creates `faluleru' for [GRAY 0.5–1.0]
 Hearer a4 adopts `faluleru' for [GRAY 0.5–1.0]
Game 100 
 Speaker a7 uses `faluleru' for [GRAY 0.5–1.0]
 Hearer a4 correctly interprets `faluleru'
\end{verbatim}
The semiotic triangles existing at this point in the population are 
summarised in \figref{triangle6}. 
Agent {\bfshape  a4} is now already in a dilemma because he has created
`notabefe' and picked up `faluleru' from {\bfshape  a7}.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj} & Hpos & Vpos & Height & Width & Gray & Area \\ \midrule
0 & 0.37 & 0.44 & 0.92 & 0.92 & 0.69 & 0.10\\ 
1 & 0.98 & 0.55 & 0.25 & 0.11 & 0.78 & 0.88\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game3}Sensory data for game 3 after scaling.}
\end{center}
\end{table}
Values for VPOS and GRAY are very close so they are not 
considered as sufficiently salient.
\begin{verbatim}
Game 3
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 2 objects: 
       square-0 rectangle-1 
  a1 chooses rectangle-1 as the topic 
  a1 considers as salient AREA WIDTH HEIGHT HPOS 
  a1 categorises the topic as [HEIGHT 0.0–0.5]
  a2 creates a new word: `mibati'
  a1 says: `mibati'
  a2 does not know `mibati'
  a2 says: `mibati?'
  a1 points to rectangle-1
  a1 considers as salient AREA WIDTH HEIGHT HPOS 
  a1 categorises the topic as [HEIGHT 0.0–0.5]
  a1 stores `mibati' as [HEIGHT 0.0–0.5]
\end{verbatim}
{}[HEIGHT 0.0–0.5] has been chosen by the speaker
because HEIGHT was one of the salient channels (even 
though clearly not the only salient one) and 
because a successful distinction already existed in 
the ontology. The same 
distinction was chosen by chance by the hearer, but he 
could just as well have chosen a distinction based on 
the AREA, WIDTH or HPOS.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap6/figs/triangle5.pdf}}
\caption{\label{triangle5}Semiotic triangles used 
in game 25. The score of the relation between [GRAY 0.0–0.5] and 
`puxazi' is increased in both agents. The relation between `puxazi' and 
{}[GRAY 0.0–0.5] gets damped.}
\end{figure}

>>>>

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape obj} & \textsc{hpos} & \textsc{vpos} & \textsc{height} & \textsc{width} & \textsc{gray} & \textsc{area} \\ \midrule
0 & 0.66 & 0.95 & 0.01 & 0.71 & 0.19 & 0.27\\ 
1 & 0.69 & 0.83 & 0.07 & 0.33 & 0.97 & 0.21\\ 
2 & 0.99 & 0.87 & 0.54 & 0.72 & 0.22 & 0.57\\ 
saliency & 0.03 & 0.05 & 0.07 & 0.39 & 0.75 & 0.06 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-rect1}Sensory data for the scene shown in \figref{rect1}.}
\end{center}
\end{table}
All rectangles are relatively close to each other and have more or less the 
same height and width. But the grayscale is clearly the more salient because rectangle-1 is 
much darker than the others. I assume that there are only two agents in the population and that they always use only 
the most salient to conceptualise the scene. 
The speaker and hearer have to traverse only 
two sides of the semiotic square (\figref{square6})
because we assume that perceived image and object being
referred to are identical for both agents.

@@

The speaker first plays a Discrimination Game traversing the 
semantic side of the square going from the referent 
{\itshape rectangle-1} to a possible meaning [\textsc{gray} 0.5–1.0]. 
He then plays a Naming Game traversing the lexical side of 
of the square to find the word `pokuneso' for this chosen meaning. 
The hearer traverses the lexical side of the triangle in 
the other direction to interpret the word `pokuneso' as
{}[\textsc{gray} 0.5–1.0], and then identifies the referent by 
filtering the objects in the context with this meaning. 
Only rectangle-1 remains, so the game succeeds. 
The whole game is reported by the commentator as follows: 
\begin{verbatim}
Game 500
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a2 chooses rectangle-1 as the topic 
  a2 categorises the topic as [GRAY 0.5–1.0]
  a2 says: pokuneso
  a1 interprets pokuneso as [GRAY 0.5–1.0]
  a1 points to rectangle-1
  a2 signals OK 
\end{verbatim}
The game is perfectly successful because both 
agents associate the word `okuneso' with 
{}[\textsc{gray} 0.5–1.0] (dark) and they perceive the scene 
in the same way.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\itshape Translation} & {\bfshape  a1}&{\bfshape  a2} \\ \midrule
{}[\textsc{hpos} 0.0–0.5] & vapola&left&-&0.1\\ 
{}[\textsc{hpos} 0.5–1.0]& gonapa&right &0.1&-\\ 
{}[\textsc{height} 0.0–0.5]&suwaxugo&short &0.6&0.8\\ 
{}[\textsc{height} 0.5–1.0]& kusone&tall &0.4&0.5\\ 
{}[\textsc{width} 0.0–0.5]&bepupepa&narrow &0.1&0.1\\ 
{}[\textsc{width} 0.0–0.25]&kutaki&very narrow &-&0.1\\ 
{}[\textsc{width} 0.5–1.0]& zikorika&wide &0.0&0.3\\ 
{}[\textsc{gray} 0.0–0.5]& fesasado&light &0.5&0.7\\ 
{}[\textsc{gray} 0.5–1.0]& pokuneso&dark &0.8&0.9\\ 
{}[\textsc{area} 0.5–1.0]& mafanoda&large &0.1&0.1\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex100}. Complete lexicon of {\bfshape  a1} and {\bfshape  a2} after 100 games.}
\end{center}
\end{table}
We see that, at this point, the agents have lexicalised 
only the most general distinctions, such as `dark' (\emph{pokuneso}) 
versus `light' (\emph{fesasado}) or `short' (\emph{suwaxugo}) versus `tall' 
(\emph{kusone}). Words for the grayscale and height dimensions
have the strongest scores, although this is purely accidental. 
When we would start another simulation from scratch,
we would end up with different words and perhaps 
other distinctions would be more successful.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\itshape Translation} & {\bfshape  a1}&{\bfshape  a2} \\ \midrule
{}[\textsc{hpos} 0.0–0.5]&vapola&left &0.7&1.0\\ 
{}[\textsc{hpos} 0.5–1.0]&gonapa&right &0.6&0.5\\ 
{}[\textsc{vpos} 0.0–0.5]&rixuzime& up & 0.2&0.7\\ 
{}[\textsc{vpos} 0.5–1.0]&gofugage& down &0.6&1.0\\ 
{}[\textsc{height} 0.0–0.5]&suwaxugo&short & 1.0&1.0\\ 
{}[\textsc{height} 0.0–0.25]&tawube&very short & 0.4&0.5\\ 
{}[\textsc{height} 0.25–0.5]&narofi&medium short&0.1&0.4\\ 
{}[\textsc{height} 0.5–1.0]&kusone&tall&1.0&1.0\\ 
{}[\textsc{height} 0.5–0.75]&wuruzo&medium tall&0.3&0.6\\ 
{}[\textsc{height} 0.75–1.0]&bowaluro&very tall&0.6&0.2\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex500a}Lexicon of {\bfshape  a1} and {\bfshape  a2} after 500 games.}
\end{center}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{}[\textsc{width} 0.0–0.5]&bepupepa&narrow & 1.0&1.0\\ 
{}[\textsc{width} 0.0–0.25]&kutaki&very narrow & 0.1&0.5\\ 
{}[\textsc{width} 0.25–0.5]&wukogo&medium narrow & 0.2&-\\ 
{}[\textsc{width} 0.5–1.0]&zikorika&wide & 1.0&1.0\\ 
{}[\textsc{width} 0.5–0.75]&mitula&medium wide &0.1&-\\ 
{}[\textsc{width} 0.75–1.0]&wupixo&very wide & -&0.2\\ 
{}[\textsc{gray} 0.0–0.5]&fesasado&light & 1.0&1.0\\ 
{}[\textsc{gray} 0.0–0.25]&sanize&very light & -&0.1\\ 
{}[\textsc{gray} 0.5–1.0]&pokuneso&dark &0.9&1.0\\ 
{}[\textsc{gray} 0.5–0.75]&wavosoru&medium dark & 0.2&0.5\\ 
{}[\textsc{gray} 0.75–1.0]&kuragoni&very dark &0.3&0.2\\ 
{}[\textsc{area} 0.0–0.5]&babifewa&small & -&0.1\\ 
{}[\textsc{area} 0.25–0.5]&togule&medium small & 0.1&0.1\\ 
{}[\textsc{area} 0.5–1.0]&mafanoda&large & 0.2&0.5\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex500b}Lexicon of {\bfshape  a1} and {\bfshape  a2} after 1000 games.}
\end{center}
\end{table}

We see that words for the basic distinctions have further established
themselves. Words for `short' (\emph{suwaxugo}) and `tall' (\emph{kusone}), or `light' (\emph{fesasado}) and `dark' (\emph{pokuneso}) now have scores of 1.0. Words for more refined categories, like `very short' (\emph{tawube}) or `very narrow' (\emph{kutaki}), are beginning to establish themselves.

Two steps in the evolution of the discrimination trees
underlying this lexicon are shown in \figref{gdis1}.
There is a progressive refinement 
of all trees as time goes on, 
because all sensory channels have the same chance
of being most salient. But the trees are not 
the same for the two agents at every stage
of development because even though they prefer to expand
the salient channel, the agents have encountered 
different environmental situations in which different
channels were salient. For example, after 100 
games, {\bfshape  a1} has less refinements for the 
\textsc{width} channel than {\bfshape  a2}. 
After 500 games, all trees 
have at least one level of refinement. Not all 
categories have been lexicalised. For example, the 
\textsc{width} channel is three levels deep in both 
agents but no words exist yet for the deepest 
level.

@@

Also for the hearer, the re-entrant flow is important. 
A hearer can only know which meaning was intended for a particular
word after trying out the meaning on the scene. He
therefore uses the context to determine the meaning of 
the utterance. For example, a particular word may mean both [\textsc{large}]
and [\textsc{dark}], particularly during the phase of 
early language acquisition. However if only one of these categories
picks out a single referent, it is chosen as 
the meaning, and the hearer will act upon this choice by 
pointing to the object it singles out from the scene.

@@

The next thing I had to do is reconsider the score
updating mechanisms, even though they are basically 
the same as used earlier (\figref{incr-decr2}). 
For a given referent, there are multiple meanings possible
(in case more than one channel is considered to 
be sufficiently salient), and
for each meaning there are multiple words. The best 
one of this whole lot is chosen by the speaker and used
for the utterance transmitted to the hearer. We have 
seen that it is important for the hearer to use lateral
inhibition based on the outcome of a game. But although 
the speaker is considering {\itshape all} the possible 
conceptualisations, lateral
inhibition should only take place between the 
lexicalisations of the meaning that was finally chosen. 
So when the game was successful, 
the speaker increases the association that was used with $\delta$
and decreases all the other associations
{\itshape with the same meaning}. \emph{delta} is still set to a reasonable
low value, namely \emph{delta} = 0.1. 
The hearer does the same as explained earlier in 
the Naming Game. The score of all
the alternative meanings for the word (or words) 
that are used by the speaker are decreased. 
When the game fails, both the speaker and the hearer 
decrease the association that they used with $\delta$. No
change takes place to the scores of any of the 
other associations.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.75\textwidth]{chap6/figs/incr-decr2.pdf}}
\caption{\label{incr-decr2}Score adjustments after a successful game. Used 
associations go up and competing associations go down.
The game producing these relationships is discussed
later as game 10008.}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj} & Hpos & Vpos & Height & Width & Gray & Area \\ \midrule
0 & 0.51 & 0.98 & 0.90 & 0.47 & 0.79 & 0.60\\ 
1 & 0.75 & 0.68 & 0.26 & 0.09 & 0.14 & 0.20\\ 
2 & 0.76 & 0.54 & 0.56 & 0.26 & 0.94 & 0.36\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game77}Sensory data for game 77 after sensor-scaling.}
\end{center}
\end{table}
The speaker is again {\bfshape  a2}. \textsc{height} is the most salient channel. 
{\bfshape  a2} has a word for [\textsc{height} 0.75–1.0] (very tall), namely
`bowaluro', and uses it. The hearer {\bfshape  a1} 
does not know the 
word, conceptualises the scene, and arrives at the 
same category, so the hearer stores the new word
with the same meaning as the speaker. 
\begin{verbatim}
Game 77
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a2 chooses rectangle-1 as the topic 
  a2 categorises the topic as [HEIGHT 0.75–1.0]
  a2 says: `bowaluro'
  a1 does not know `bowaluro'
  a1 says: `bowaluro?'
  a2 points to rectangle-1
  a1 categorises the topic as [HEIGHT 0.75–1.0]
  a1 stores `bowaluro' as [HEIGHT 0.75–1.0]
\end{verbatim}
The reason why {\bfshape  a1} has guessed the right 
meaning of `bowaluro' is because both agents use
only the most salient channel and they both share
the same perception of reality. We will soon see 
that if these constraints are not valid, agents 
are not always so lucky and hence multiple meanings
start to circulate for the same word.

@@

Part of the lexicons of five of the ten agents after 2000 games 
(only those words which have positive scores) are shown in the 
\tabref{tab:lex2000}. We see that synonymy does indeed occur. For
example, two words are in the 
running for [\textsc{hpos} 0.5–1.0]: `rutaxese' and `xomupovi', and 
three for [\textsc{vpos} 0.5–1.0]: `wavone', `zaxawe', and 
`dazofo'. Some words are already well established, for
example `numefuli' for [\textsc{gray} 0.5–1.0] (dark).

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l  }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\bfshape  a1}&{\bfshape  a2}&{\bfshape  a3}&{\bfshape  a4}&{\bfshape  a5} \\ \midrule
{}[\textsc{hpos} 0.5–1.0]&rutaxese& &0.1& & &\\ 
 & xomupovi&0.1& &0.2& &0.1\\ 
{}[\textsc{vpos} 0.5–1.0]&wavone& & &0.3& &\\ 
 & zaxawe& & & &0.1& \\ 
 & dazofo&0.1& & &0.2&\\ 
{}[\textsc{width} 0.0–0.5]&buxevo& & & & & \\ 
 & vubupo&1.0&0.6&1.0&0.1&1.0\\ 
{}[\textsc{width} 0.5–1.0]&pawixona& & &0.1& & \\ 
 & rikepule&1.0&0.8&1.0&1.0&1.0\\ 
{}[\textsc{width} 0.5–0.75]&gowinoge& & &0.1& &  \\ 
 & wesurodi& & & &0.2&\\ 
{}[\textsc{width} 0.75–1.0]&besabi& & &0.2& & \\ 
 & lituvi&0.1& & & & \\ 
{}[\textsc{gray} 0.5–0.75]&goxomixe& &0.1& & & \\ 
 & korufo& & & &0.2&\\ 
{}[\textsc{gray} 0.0–0.5]&numefuli&1.0&0.7&1.0&1.0&1.0\\ 
{}[\textsc{gray} 0.0–0.25]&rekemaxi&0.2& & & & \\ 
{}[\textsc{gray} 0.5–1.0]&faluleru&0.6&0.6&1.0&1.0&1.0\\ 
 & nupanu& & & & &0.2 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex2000}Lexicon of five agents after 2000 games.}
\end{center}
\end{table}

The positive feedback loop between use and success
has already dampened some synonyms. For some cases, like [\textsc{gray} 0.5–1.0] the competition has died out 
with one word `faluleru' being the winner. 
For others, like [\textsc{hpos} 0.5–1.0], the competition 
is still going on, although we can guess
that `xomupovi' is probably going to be the winner.

It is instructive to follow the history of the words 
in use for a particular meaning. For 
example, let us look at the words for [\textsc{gray} 0.5–1.0] 
(dark) in the very early phases of lexicon development.
Three different words are quickly created, and 
`faluleru' is the first one that has some success. 
\begin{verbatim}
Game 76
 Speaker a4 creates `notabefe' for [GRAY 0.5–1.0]
 Hearer a5 adopts `notabefe' for [GRAY 0.5–1.0]
Game 79 
 Speaker a2 creates `vivevobo' for [GRAY 0.5–1.0]
 Hearer a7 does not adopt `vivebo' 
     (failed to discriminate)
Game 88 
 Speaker a7 creates `faluleru' for [GRAY 0.5–1.0]
 Hearer a4 adopts `faluleru' for [GRAY 0.5–1.0]
Game 100 
 Speaker a7 uses `faluleru' for [GRAY 0.5–1.0]
 Hearer a4 correctly interprets `faluleru'
\end{verbatim}
The semiotic triangles existing at this point in the population are 
summarised in \figref{triangle6}. 
Agent {\bfshape  a4} is now already in a dilemma because he has created
`notabefe' and picked up `faluleru' from {\bfshape  a7}.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj} & Hpos & Vpos & Height & Width & Gray & Area \\ \midrule
0 & 0.37 & 0.44 & 0.92 & 0.92 & 0.69 & 0.10\\ 
1 & 0.98 & 0.55 & 0.25 & 0.11 & 0.78 & 0.88\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game3}Sensory data for game 3 after scaling.}
\end{center}
\end{table}
Values for \textsc{vpos} and \textsc{gray} are very close so they are not 
considered as sufficiently salient.
\begin{verbatim}
Game 3
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 2 objects: 
       square-0 rectangle-1 
  a1 chooses rectangle-1 as the topic 
  a1 considers as salient AREA WIDTH HEIGHT HPOS 
  a1 categorises the topic as [HEIGHT 0.0–0.5]
  a2 creates a new word: `mibati'
  a1 says: `mibati'
  a2 does not know `mibati'
  a2 says: `mibati?'
  a1 points to rectangle-1
  a1 considers as salient AREA WIDTH HEIGHT HPOS 
  a1 categorises the topic as [HEIGHT 0.0–0.5]
  a1 stores `mibati' as [HEIGHT 0.0–0.5]
\end{verbatim}
{}[\textsc{height} 0.0–0.5] has been chosen by the speaker
because \textsc{height} was one of the salient channels (even 
though clearly not the only salient one) and 
because a successful distinction already existed in 
the ontology. The same 
distinction was chosen by chance by the hearer, but he 
could just as well have chosen a distinction based on 
the \textsc{area}, \textsc{width} or \textsc{hpos}.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap6/figs/triangle5.pdf}}
\caption{\label{triangle5}Semiotic triangles used 
in game 25. The score of the relation between [\textsc{gray} 0.0–0.5] and 
`puxazi' is increased in both agents. The relation between `puxazi' and 
{}[\textsc{gray} 0.0–0.5] gets damped.}
\end{figure}
